# Data visualisation

```{r setup}
library(tidyverse)
```

## Introduction
 
## First steps

### Content 
```{r}
data(mpg)
```

Find out about the correlation between `displ` (= a car's engine size in liters) and `hwy` (= a car's fuel efficiency on the highway in miles per gallon):

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

### Exercises
1.  Run `ggplot(data = mpg)` what do you see?
```{r}
ggplot(data = mpg)
```
Nothing because the geometric function and mappings are missing.

1.  How many rows are in `mtcars`? How many columns?
```{r}
data(mtcars)
nrow(mtcars) 
ncol(mtcars) 
```

1.  What does the `drv` variable describe?  Read the help for `?mpg` to find
    out.
```{r, eval = F}
?mpg
```
f = front-wheel drive; r = rear-wheel drive; 4 = four-wheel drive
     
1.  Make a scatterplot of `hwy` vs `cyl`.
```{r}
ggplot(data = mpg) +
  geom_point(aes(x = hwy, y = cyl))
```

1.  What happens if you make a scatterplot of `class` vs `drv`. Why is
    the plot not useful?
```{r}
ggplot(mpg) +
  geom_point(aes(x = class, y = drv))
```
There are multiple y-values for one x-value. Also, class is not a numerical variable, but a categorial one.

## Aesthetic mappings

## Common problems

## Facets

## Geometric objects

## Statistical transformations

## Position adjustments

## Coordinate systems

## The layered grammar of graphics

