---
title: "Chapter 03 Data Visualization"
author: "kaetschap"
date: "November 23, 2016"
output: html_document
---

# Data visualisation

```{r, message = F}
library(tidyverse)
```

## Introduction
 
## First steps

### Content 

```{r}
data(mpg)
```

Find out about the correlation between `displ` (= a car's engine size in liters) and `hwy` (= a car's fuel efficiency on the highway in miles per gallon):

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

### Exercises
1.  Run `ggplot(data = mpg)` what do you see? 
    
    ```{r}
    ggplot(data = mpg)
    ```
    
    Nothing because the geometric function and mappings are missing.

1.  How many rows are in `mtcars`? How many columns?
    
    ```{r}
    data(mtcars)
    nrow(mtcars) 
    ncol(mtcars) 
    ```

1.  What does the `drv` variable describe?  Read the help for `?mpg` to find
    out.
    
    ```{r, eval = F}
    ?mpg
    ```
    
    f = front-wheel drive; r = rear-wheel drive; 4 = four-wheel drive
     
1.  Make a scatterplot of `hwy` vs `cyl`.
    
    ```{r}
    ggplot(data = mpg) +
      geom_point(aes(x = hwy, y = cyl))
    ```

1.  What happens if you make a scatterplot of `class` vs `drv`. Why is
    the plot not useful?
    
    ```{r}
    ggplot(mpg) +
      geom_point(aes(x = class, y = drv))
    ```
    
    There are multiple y-values for one x-value. Also, class is not a numerical
    variable, but a categorial one.

## Aesthetic mappings

### Content
use another aesthetic to add a third variable, class, and 
use color at unique levels to differenciate each unique value 
of the class-variable

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

other possible aethetics:

*    size  (=> size of the points)
*    alpha (=> opacity)
*    shape (=> shapes, only six possible ones; rest will stay unplotted)

manually setting the aesthetic:
since setting an aesthetic to a constant is independent of any variable, set this **outside** of the aes()-function

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

### Exercises

1.  What's gone wrong with this code? Why are the points not blue?

    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
    ```
    
    It's not blue because the color setting was used inside aes()
    The correct way:
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
    ```

1.  Which variables in `mpg` are categorical? Which variables are continuous? 
    (Hint: type `?mpg` to read the documentation for the dataset). How
    can you see this information when you run `mpg`?
    
    ```{r}
    lapply(mpg, class)
    ```
    
    continuous variables:
    ```{r}
    colnames(mpg)[grep("numeric", lapply(mpg, class))]
    ```
    
    categorial variables:
    ```{r}
    colnames(mpg)[grep("character", lapply(mpg, class))]
    ```
    
    discrete variables:
    ```{r}
    colnames(mpg)[grep("integer", lapply(mpg, class))]    
    ```

1.  Map a continuous variable to `color`, `size`, and `shape`. How do
    these aesthetics behave differently for categorical vs. continuous
    variables? 
    
    `color`:
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, color = displ))
    ```
    
    `size`:
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, size = displ))
    ```
    
    `shape`:
    ```{r, eval = F}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, shape = displ))
    ```
    `size` & `color` work well, `shape` is more suited for categorial variables
    
    `color`:
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, color = drv))
    ```
    
    `size`:
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, size = drv))
    ```
    
    `shape`:
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, shape = drv))
    ```
    
    `color` and `shape` work well, `size` is better for discrete/continuous
    variables
    
1.  What happens if you map the same variable to multiple aesthetics? 
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, color = hwy))
    ```
    
    The plot essentially loses information, since the same thing is displayed
    twice

1.  What does the `stroke` aesthetic do? What shapes does it work with?
    (Hint: use `?geom_point`)
    
    ```{r, eval = F}
    ?geom_point
    ```
    
    > For shapes that have a border (like 21), you can colour the inside and
    > outside separately. Use the stroke aesthetic to modify the width of the
    > border
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, stroke = 2))
    ```
    
1.  What happens if you map an aesthetic to something other than a variable 
    name, like `aes(colour = displ < 5)`? 
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, color = displ < 5))
    ```
    
    This creates a logical vector indicating whether the condition is true or
    not and displays these two levels as the third aesthetic in the legend

## Common problems

## Facets
### Content
### Exercises

1.  What happens if you facet on a continuous variable?

    ```{r}
    ggplot(data = mpg) +
      geom_point(aes(x = drv, y = cyl)) +
      facet_wrap(~ displ, nrow = 2)
    ```
  
    For every realisation of the variable a new facet will be added.
    
1.  What do the empty cells in plot with `facet_grid(drv ~ cyl)` mean?
    How do they relate to this plot?
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = drv, y = cyl))
    ```
    
    There are points missing whenever there is no realisation in the data as
    `cyl` for that `drv`.

1.  What plots does the following code make? What does `.` do?

    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy)) +
      facet_grid(drv ~ .)
    
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy)) +
      facet_grid(. ~ cyl)
    ```
    
    The `facet_grid` formula works according to the pattern `y ~ x`, where the 
    y variable is mapped along the y-axis of the plot (as rows) and the x
    variable is mapped along the x-axis (as columns).
    
    When instead of a variable, there is a dot, that means to omit that
    dimension. (`. ~ x` to map `x` along columns and `y ~ .` to map `y` along
    rows).

1.  Take the first faceted plot in this section:

    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy)) + 
      facet_wrap(~ class, nrow = 2)
    ```
    
    What are the advantages to using faceting instead of the colour aesthetic?
    What are the disadvantages? How might the balance change if you had a 
    larger dataset?
    
    *   advantages:
        Faceting gives a plot per level. That means that it's easier to spot
        how the data for each level is distributed within the x- and y-axes.
        This is especially helpful for larger datasets, where the data points 
        from different facets might overlap.
        
    *   disadvantages:
        Since levels are split up over multiple plots, it is harder to spot 
        outliers or clusters in the data.
    
1.  Read `?facet_wrap`. What does `nrow` do? What does `ncol` do? What other
    options control the layout of the individual panels? Why doesn't
    `facet_grid()` have `nrow` and `ncol` variables?
    
    `facet_wrap()` builds facetting for one vector. `nrow` and `ncol` set the
    number of rows and columns along which the facets of that vector should be 
    distributed in the plot.
    
    `facet_grid()` has predetermined numbers of rows and columns. The number of
    rows is equal to the number of levels for the `y` variable and the number 
    of columns to the number of level for the `x` variable specified in the 
    formula `y ~ x`.

1.  When using `facet_grid()` you should usually put the variable with more
    unique levels in the columns. Why?
    
    This makes sense because most displaying devices have a larger number of 
    pixels on the horizontal than the vertical axis.

## Geometric objects

### Content

### Exercises

1.  What geom would you use to draw a line chart? A boxplot? 
    A histogram? An area chart?
    
    *   line: 
        `geom_line()`
        
    *   boxplot: 
        `geom_boxplot()`
        
    *   histogram:
        `geom_histogram()`
        
    *   area chart:
        `geom_area()`
        
        
1.  Run this code in your head and predict what the output will look like.
    Then, run the code in R and check your predictions.
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
      geom_point() + 
      geom_smooth(se = FALSE)
    ```
    
    This plot will show a scatter plot of `hwy` against `displ` with an 
    additional aesthetic for `drv` by color. On top, there will be a smoothed 
    regression line per level of `drv` without a confidence interval being 
    displayed.

1.  What does `show.legend = FALSE` do?  What happens if you remove it?  
    Why do you think I used it earlier in the chapter?
    
    It didn't occur earlier in the chapter...
    But it stops the legend from being displayed.

1.  What does the `se` argument to `geom_smooth()` do?
    
    `se` shows whether a confidence interval should be displayed around the 
    smoothed curve.

1.  Will these two graphs look different? Why/why not?

    ```{r, eval = FALSE}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point() + 
      geom_smooth()
    
    ggplot() + 
      geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
    ```
    
    Yes, they will. In the first, options are set in the `ggplot()`-function 
    and passed down to the next layers. In the second plot, the options are set
    manually per layer to the same values.

1.  Recreate the R code necessary to generate the following graphs.
    
    ```{r echo = FALSE, fig.width = 3, out.width = "50%", fig.align = "default", message = FALSE}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point() + 
      geom_smooth(se = FALSE)
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_smooth(aes(group = drv), se = FALSE) +
      geom_point()
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
      geom_point() + 
      geom_smooth(se = FALSE)
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point(aes(color = drv)) + 
      geom_smooth(se = FALSE)
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point(aes(color = drv)) +
      geom_smooth(aes(linetype = drv), se = FALSE)
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point(size = 4, colour = "white") + 
      geom_point(aes(colour = drv))
    ```
    
    ```{r}
    ggplot(mpg, aes(x = displ, y = hwy)) + 
      geom_point() +
      geom_smooth(se = F)
    
    ggplot(mpg, aes(x = displ, y = hwy, group = drv)) +
      geom_point() +
      geom_smooth(se = F)
    
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
      geom_point() + 
      geom_smooth(se = FALSE)
    
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point(aes(color = drv)) + 
      geom_smooth(se = FALSE)
    
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point(aes(color = drv)) + 
      geom_smooth(se = FALSE, aes(linetype = drv))
    
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
      geom_point(stroke = 2, color = "white") +
      geom_point()
    ```

## Statistical transformations

### Content
It is possible to use geoms and stats interchangeably. Each geom has its default stat, and each stat its default geom. For example, `geom_boxplot` uses `stat_count` and vice versa.

Specify the stat in the plotting function like this:

```{r, eval=F}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = a, y = b), stat = "identity")
```

### Exercises

1.  What is the default geom associated with `stat_summary()`? How could
    you rewrite the previous plot to use that geom function instead of the 
    stat function?
    
    The `stat_summary()` function uses `geom = "pointrange"` by default.
    
    Previous plot:
    ```{r}
    ggplot(diamonds, aes(x = cut, y = depth)) + 
      stat_summary(
        fun.ymin = min,
        fun.ymax = max,
        fun.y = median
      )
    ```
    
    
    ```{r}
    diamonds %>% 
      select(cut, depth) %>%
      group_by(cut) %>%
      summarize(mindepth = min(depth), maxdepth = max(depth), mediandepth = median(depth)) %>% 
      ggplot(aes(x = cut, y = mediandepth, ymin = mindepth, ymax = maxdepth)) + 
      geom_pointrange()
    ```

1.  What does `geom_col()` do? How is it different to `geom_bar()`?

    ```{r, eval = F}
    ?geom_bar()
    ```
    
    > `geom_bar` makes the height of the bar proportional to the number of cases 
    > in each group […]. If you want the heights of the bars to represent values
    > in the data, use `geom_col` instead. `geom_bar` uses `stat_count` by 
    > default […]. `geom_col` uses `stat_identity`.

1.  Most geoms and stats come in pairs that are almost always used in 
    concert. Read through the documentation and make a list of all the 
    pairs. What do they have in common?
    
    http://docs.ggplot2.org/current/index.html: see stats for each geom

1.  What variables does `stat_smooth()` compute? What parameters control
    its behaviour?
    
    `stat_smooth` uses the same arguments as `geom_smooth` but can be used 
    whenever a non-standard geom should be plotted.
    It computes a regression line and standard error bounds.
    It is possible to specify:
    *   `method`: the smoothing method
    *   `formula`: the regression formula to be used
    *   `se`: whether to display a confidence interval around the smoothed 
    regression line
    *   etc (read documentation)

1.  In our proportion bar chart, we need to set `group = 1`. Why? In other
    words what is the problem with these two graphs?
    
    ```{r, eval = FALSE}
    ggplot(data = diamonds) + 
      geom_bar(mapping = aes(x = cut, y = ..prop..))
    ggplot(data = diamonds) + 
      geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))
    ```
    
    `group = 1` specifies that all values should be put into one grouping. This 
    is necessary for the values to be in proportion to the whole population as
    opposed to just their own group.
    Without grouping, R computes the values in the plot as follows:
    fair is 100% of fair, good is 100% of good
    If the option is set, R computes them like this:
    fair is 3% of the whole sample, good is 9% etc.

## Position adjustments

### Content

### Exercises

1.  What is the problem with this plot? How could you improve it?

    ```{r}
    ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
      geom_point()
    ```
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
      geom_point(position = "jitter")
    ```
    
    In order to get a feeling for how many points cluster at each spot it is 
    helpful to either use jitter, which moves the points by a random amount in 
    height and width, or opacity.

1.  What parameters to `geom_jitter()` control the amount of jittering?

    `width` and `height` and be set to percentage values with a default of 40% 
    each.

1.  Compare and contrast `geom_jitter()` with `geom_count()`.
    ```{r}
    ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
    geom_count()
    ```
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
    geom_jitter()
    ```
    
    `geom_count()` counts the number of observations that occur at each point in 
    the coordinate system. The higher that number, the bigger the point in the 
    point chart.
    
    `geom_jitter()` moves the points instead, so that areas with more 
    observations are visibly denser than others.
    
1.  What's the default position adjustment for `geom_boxplot()`? Create
    a visualisation of the `mpg` dataset that demonstrates it.

    The default setting for `geom_boxplot()` is `position = "dodge"`.
    
    ```{r}
    ggplot(mpg, aes(x = class, y = hwy)) +
      geom_boxplot()
    ```
    
    This means that next to each boxplot, there is some space to the side until 
    the next one follows.
    
## Coordinate systems

### Content

### Exercises

1.  Turn a stacked bar chart into a pie chart using `coord_polar()`.

1.  What does `labs()` do? Read the documentation.

1.  What's the difference between `coord_quickmap()` and `coord_map()`?

1.  What does the plot below tell you about the relationship between city
    and highway mpg? Why is `coord_fixed()` important? What does 
    `geom_abline()` do?
    
    ```{r, fig.asp = 1, out.width = "50%"}
    ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
      geom_point() + 
      geom_abline() +
      coord_fixed()
    ```

## The layered grammar of graphics

