---
title: "Chapter_05_DataTransformation"
author: "kaetschap"
date: "15 Dezember 2016"
output: html_document
---

```{r, message=F, warning=F}
library(tidyverse)
library(nycflights13)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, indent = '    ')
```

# Exercises: `filter()`

1.  Find all flights that

    1. Had an arrival delay of two or more hours
    
    ```{r}
    flights %>% filter(arr_delay >= 120)
    ```
       
    1. Flew to Houston (`IAH` or `HOU`)
    
    ```{r}
    flights %>% filter(dest %in% c('IAH', 'HOU'))
    ```
    
    1. Were operated by United, American, or Delta
    
    ```{r}
    flights %>% filter(carrier %in% c('UA', 'AA', 'DL'))
    ```
    
    1. Departed in summer (July, August, and September)
    
    ```{r}
    flights %>% filter(month %in% c(7,8,9))
    ```
    
    1. Arrived more than two hours late, but didn't leave late
    
    ```{r}
    flights %>% filter(dep_delay <= 0 & arr_delay > 120)
    ```
    
    1. Were delayed by at least an hour, but made up over 30 minutes in flight
    
    ```{r}
    flights %>% filter(dep_delay >= 60 & arr_delay < (dep_delay - 30))
    ```
    
    1. Departed between midnight and 6am (inclusive)
    
    ```{r}
    flights %>% filter(dep_time >= 0 & dep_time <= 600)
    ```


1.  Another useful dplyr filtering helper is `between()`. What does it do?
    Can you use it to simplify the code needed to answer the previous 
    challenges?
    ```{r}
    flights %>% filter( between(dep_time, 0, 600) )
    ```

1.  How many flights have a missing `dep_time`? What other variables are 
    missing? What might these rows represent?
    ```{r}
    flights %>% filter(is.na(dep_time)) %>% nrow
    ```
        
    ```{r, message = F}
    cols_numberna <- lapply(flights, function(x) {
        y <- is.na(x)
        r <- length(which(y == TRUE))
    })
    ```
    
    ```{r}
    colnames(flights)[which(lapply(cols_numberna, function(x) x > 0) == TRUE)]
    ```

1.  Why is `NA ^ 0` not missing? Why is `NA | TRUE` not missing?
    Why is `FALSE & NA` not missing? Can you figure out the general
    rule?  (`NA * 0` is a tricky counterexample!)

    `NA | TRUE` ≠ `NA` since if you want to filter on two conditions joined by "OR", and one already evaluates to `TRUE`, the `NA` value is no longer relevant for the result. The result will be `TRUE` independent of whether the `NA` condition is a missing `TRUE` or a missing `FALSE`.
    
    `NA & FALSE` ≠ `NA` since if you want to filter on two conditions joined by "AND", and one already evaluates to `FALSE`, the `NA` value is no longer relevant for the result. The result will be `FALSE` independent of whether the `NA` condition is a missing `TRUE` or a missing `FALSE`.
    
    `NA ^ 0` ≠ `NA` since taking any number to the power of `0` will yield `1`. Thus, it does not matter which number the `NA` value stands for, and it is safe to say the result will be `1`.
    
    So by and large, if a computation will always yield the same result, regardless of the value of the argument, a `NA` value will not come in the way of a correct computation. The exception to this is `NA * 0` which evaluates to `NA`.
    
# Exercises: `arrange()`

1.  How could you use `arrange()` to sort all missing values to the start?
    (Hint: use `is.na()`).
    ```{r}
    flights %>% arrange(!is.na(dep_time))
    ```
    
1.  Sort `flights` to find the most delayed flights. Find the flights that
    left earliest.
    ```{r}
    flights %>% arrange(desc(arr_delay)) %>% head(5)
    flights %>% arrange(dep_time) %>% head(5)
    ```

1.  Sort `flights` to find the fastest flights.

    Fast in terms of air time:
    
    ```{r}
    flights %>% 
        select(origin, dest, air_time) %>% 
        filter(!is.na(air_time)) %>% 
        group_by(origin, dest) %>% 
        summarize(min_air_time = min(air_time), max_air_time = max(air_time), mean_air_time = round(mean(air_time), 2)) %>%
        arrange(min_air_time)
    ```
    
    Fast in terms of smallest delay:
    ```{r}
    flights %>% arrange((dep_delay+arr_delay))
    ```

1.  Which flights travelled the longest? Which travelled the shortest?
    ```{r}
    flights %>% arrange(desc(air_time)) %>% head(5)
    flights %>% arrange(air_time) %>% head(5)
    ```

#Exercises: `select()`

1.  Brainstorm as many ways as possible to select `dep_time`, `dep_delay`,
    `arr_time`, and `arr_delay` from `flights`.
    
    No.
    
1.  What happens if you include the name of a variable multiple times in
    a `select()` call?
    
    The column appears only once, at the first location it was mentioned.
    ```{r}
    flights %>% select(air_time, distance, air_time)
    ```
  
1.  What does the `one_of()` function do? Why might it be helpful in conjunction
    with this vector?
    
    ```{r}
    vars <- c("year", "month", "day", "dep_delay", "arr_delay")
    ```

    This fails:
    ```{r, eval = F}
    flights %>% select(vars)
    ```
    
    `one_of()` makes it possible to select the columns mentioned in the vector:
    
    ```{r, eval = F}
    flights %>% select(one_of(vars))
    ```
    
    It is equivalent to:
    ```{r, eval = F}
    flights %>% select_(.dots = lapply(vars, as.symbol))
    ```
    
    
1.  Does the result of running the following code surprise you?  How do the
    select helpers deal with case by default? How can you change that default?

    ```{r, eval = FALSE}
    select(flights, contains("TIME"))
    ```
    
    Yes, since operations in base-R are case-sensitive by default.
    
    ```{r, eval = F}
    select(flights, contains("TIME", ignore.case = FALSE))
    ```

# Exercises: `mutate()`

```{r, eval = FALSE, echo = FALSE}
flights <- flights %>% mutate(
  dep_time = hour * 60 + minute,
  arr_time = (arr_time %/% 100) * 60 + (arr_time %% 100),
  airtime2 = arr_time - dep_time,
  dep_sched = dep_time + dep_delay
)

ggplot(flights, aes(dep_sched)) + geom_histogram(binwidth = 60)
ggplot(flights, aes(dep_sched %% 60)) + geom_histogram(binwidth = 1)
ggplot(flights, aes(air_time - airtime2)) + geom_histogram()
```

1.  Currently `dep_time` and `sched_dep_time` are convenient to look at, but
    hard to compute with because they're not really continuous numbers. 
    Convert them to a more convenient representation of number of minutes
    since midnight.
    
    ```{r}
    flights2 <- 
      flights %>% 
      mutate(
        arr_time = (arr_time %/% 100) * 60 + (arr_time %% 100),
        sched_arr_time = (sched_arr_time %/% 100) * 60 + (sched_arr_time %% 100),
        dep_time = (dep_time %/% 100) * 60 + (dep_time %% 100),
        sched_dep_time = (sched_dep_time %/% 100) * 60 + (sched_dep_time %% 100)
      )
    ```
    
1.  Compare `air_time` with `arr_time - dep_time`. What do you expect to see?
    What do you see? What do you need to do to fix it?
    
    ```{r}
    flights2 %>% 
      select(flight, arr_time, dep_time, air_time) %>%
      mutate(air_time2 = (arr_time - dep_time))
    ```
    
    
1.  Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you
    expect those three numbers to be related?
    
    ```{r}
    flights2 %>%
      select(dep_time, sched_dep_time, dep_delay) %>%
      mutate(dep_time2 = sched_dep_time + dep_delay) %>%
      mutate(dep_dev = dep_time - dep_time2) %>%
      filter(dep_dev > 0)
    ```

1.  Find the 10 most delayed flights using a ranking function. How do you want 
    to handle ties? Carefully read the documentation for `min_rank()`.
    
    ```{r}
    flights2 %>%
      select(flight, sched_dep_time, dep_time, dep_delay) %>%
      mutate(rank_dep_delay = min_rank(dep_delay)) %>%
      arrange(rank_dep_delay) %>%
      head(10)
    ```

1.  What does `1:3 + 1:10` return? Why?

    It returns a warning message: `Warning message: In c(1:3) + c(1:10) : longer object length is not a multiple of shorter object 
    length`. This is since 10 is not a multiple of three, else R would have taken the shorter vector multiple times end-to-end, like in 
    this example:
    
    ```{r}
    1:5
    1:10
    1:5 + 1:10
    ```
    

1.  What trigonometric functions does R provide?

    ```{r, eval = F}
    ??trigono
    ```
    
    R provides trigonometric functions via `base::Trig`, namely `cos()`, `sin()`, `tan()`, `acos()`, `asin()`, `atan()`, `atan2()`,   
    `cospi()`, `sinpi()`, `tanpi()`.
    

# Exercises: `summarize()`

1.  Brainstorm at least 5 different ways to assess the typical delay 
    characteristics of a group of flights. Consider the following scenarios:
    
    * A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of 
      the time.
      
    * A flight is always 10 minutes late.

    * A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of 
      the time.
      
    * 99% of the time a flight is on time. 1% of the time it's 2 hours late.
    
    Which is more important: arrival delay or departure delay?
    
    Departure delay could still be made up in the air, by flying faster for example. Arrival delay is what counts.
    
    ```{r}
    not_cancelled <- flights %>% filter(!is.na(dep_delay), !is.na(arr_delay))
    
    not_cancelled %>% summarize(med_arr_delay = median(arr_delay), min_arr_delay = min(arr_delay), max_arr_delay = max(arr_delay))

    ```

1.  Come up with another approach that will give you the same output as 
    `not_cancelled %>% count(dest)` and 
    `not_cancelled %>% count(tailnum, wt = distance)` (without using 
    `count()`).
    
    ```{r}
    not_cancelled %>% group_by(dest) %>% summarize(delay = mean(arr_delay), n = n()) %>% select(-delay)
    not_cancelled %>% group_by(tailnum) %>% summarize(n = sum(distance))
    ```

1.  Our definition of cancelled flights (`is.na(dep_delay) | is.na(arr_delay)`
    ) is slightly suboptimal. Why? Which is the most important column?
    
    ```{r}
    flights %>% filter(is.na(dep_delay) & !is.na(arr_delay)) %>% nrow
    flights %>% filter(!is.na(dep_delay) & is.na(arr_delay)) %>% nrow
    flights %>% filter(is.na(arr_delay)) %>% nrow
    flights %>% filter(is.na(air_time)) %>% nrow
    ```
    
    There are no flight with an departure delay of `NA` and a non-missing arrival delay. The other way around, there are 1,175 flights. 
    Thus, the arrival delay is more important. However, whenever the arrival delay is `NA`, so is air time. So this column may be the 
    better indicator of a cancelled flight.

1.  Look at the number of cancelled flights per day. Is there a pattern?
    Is the proportion of cancelled flights related to the average delay?
    
    ```{r}
    cancelled <- flights %>% filter(is.na(air_time))
    cancelled_ymd <- cancelled %>% mutate(date = as.Date(paste0(year, "-",month,"-",day))) %>% count(date) 
    
    ggplot(cancelled_ymd, aes(x = date, y = n)) +
      geom_line()
    ```

1.  Which carrier has the worst delays? Challenge: can you disentangle the
    effects of bad airports vs. bad carriers? Why/why not? (Hint: think about
    `flights %>% group_by(carrier, dest) %>% summarise(n())`)
    
    ```{r}
    flights %>% 
      filter(!is.na(arr_delay)) %>%
      group_by(carrier) %>%
      summarize(avg_delay = mean(arr_delay)) %>%
      arrange(desc(avg_delay))
    ```
    
    ```{r}
    airlines 
    ```
    
    F9 (Frontier Airlines) has the worst delays. 
    
    If bad carriers have a high share of flights in an airport, they are going to affect that airport's delays. If an airline has a high share of bad airports, they are going to make the airline look bad. It is necessary to look at the share of airports and airlines - if airlines are relatively evenly distributed amonst airports and one airline uses each airport around the same number of time, it should be possible to make the distinction between bad airports and bad airlines.

1.  For each plane, count the number of flights before the first delay 
    of greater than 1 hour.
    
    ```{r}
    flights %>% 
      group_by(tailnum) %>% 
      mutate(arr_date = as.Date(paste(year, month, day, sep = "-"), "%Y-%m-%d")) %>%
      arrange(arr_date) %>%
      mutate(flights_first_delay_and_after = cumany(arr_delay > 60)) %>%
      filter(!flights_first_delay_and_after) %>%
      tally(sort = T)
    ```

1.  What does the `sort` argument to `count()` do. When might you use it?

    ```{r}
    flights %>%
      count(tailnum, sort = T)
    ```
    
    The `sort` argument causes the output table to be sorted by number of occurences in descending order. In the above example this is useful because it displays the planes with the highest numbers of flights first, instead of sorting alphabetically.

# Exercises: `group_by()` and `mutate()`

1.  Refer back to the table of useful mutate and filtering functions. 
    Describe how each operation changes when you combine it with grouping.
    
    ```{r}

    ```

1.  Which plane (`tailnum`) has the worst on-time record?

    ```{r}
    flights %>%
      group_by(tailnum) %>%
      summarize(mean_arr_delay = mean(arr_delay, na.rm = T)) %>%
      filter(mean_arr_delay > 0) %>%
      arrange(desc(mean_arr_delay))
    ```
    
    N844MH has an average arrival delay of 320 minutes (5 hours and 20 minutes).

1.  What time of day should you fly if you want to avoid delays as much
    as possible?
    
    ```{r}
    flights %>%
      group_by(hour) %>%
      summarize(mean_arr_delay = mean(arr_delay, na.rm = T)) %>%
      arrange(mean_arr_delay)
    ```
    
    The smallest delays occur between 7:00 and 7:59, where flights are over five minutes early to arrive. To avoid delays it is sufficient to fly between 5 and 9:59 am.
    
1.  For each destination, compute the total minutes of delay. For each, 
    flight, compute the proportion of the total delay for its destination.
    
    ```{r}

    ```
    
1.  Delays are typically temporally correlated: even once the problem that
    caused the initial delay has been resolved, later flights are delayed 
    to allow earlier flights to leave. Using `lag()` explore how the delay
    of a flight is related to the delay of the immediately preceding flight.
    
    ```{r}

    ```
    
1.  Look at each destination. Can you find flights that are suspiciously
    fast? (i.e. flights that represent a potential data entry error). Compute
    the air time a flight relative to the shortest flight to that destination.
    Which flights were most delayed in the air?
    
    ```{r}

    ```
    
1.  Find all destinations that are flown by at least two carriers. Use that
    information to rank the carriers.
    
    ```{r}

    ```
