---
title: "Chapter_04_WorkflowBasics"
author: "kaetschap"
date: "November 26, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Exercises

1.  Why does this code not work?

    ```{r, error = TRUE}
    my_variable <- 10
    my_varıable
    ```
    
    Look carefully! (This may seem like an exercise in pointlessness, but
    training your brain to notice even the tiniest difference will pay off
    when programming.)
    
    It should be called `my_variable`, not `my_varıable`.
    
1.  Tweak each of the following R commands so that they run correctly:

    ```{r, eval = FALSE}
    library(tidyverse)

    ggplot(dota = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy))
    
    fliter(mpg, cyl = 8)
    filter(diamond, carat > 3)
    ```
    
    
    ```{r, eval = F}
    library(tidyverse)
    
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy))
    
    filter(mpg, cyl = 8)
    filter(diamonds, carat > 3)
    ```
    
1.  Press Alt + Shift + K. What happens? How can you get to the same place
    using the menus?
    
    A keyboard shorcut quickreference guide for Rstudio opens up. Alternatively, go to Tools ⇒ 
    Keyboard Shortcuts Help
