---
title: "Chapter 03 Data Visualization"
author: "kaetschap"
date: "November 23, 2016"
output: html_document
---

# Data visualisation

```{r, message = F}
library(tidyverse)
```

## Introduction
 
## First steps

### Content 

```{r}
data(mpg)
```

Find out about the correlation between `displ` (= a car's engine size in liters) and `hwy` (= a car's fuel efficiency on the highway in miles per gallon):

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

### Exercises
1.  Run `ggplot(data = mpg)` what do you see? 
    
    ```{r}
    ggplot(data = mpg)
    ```
    
    Nothing because the geometric function and mappings are missing.

1.  How many rows are in `mtcars`? How many columns?
    
    ```{r}
    data(mtcars)
    nrow(mtcars) 
    ncol(mtcars) 
    ```

1.  What does the `drv` variable describe?  Read the help for `?mpg` to find
    out.
    
    ```{r, eval = F}
    ?mpg
    ```
    
    f = front-wheel drive; r = rear-wheel drive; 4 = four-wheel drive
     
1.  Make a scatterplot of `hwy` vs `cyl`.
    
    ```{r}
    ggplot(data = mpg) +
      geom_point(aes(x = hwy, y = cyl))
    ```

1.  What happens if you make a scatterplot of `class` vs `drv`. Why is
    the plot not useful?
    
    ```{r}
    ggplot(mpg) +
      geom_point(aes(x = class, y = drv))
    ```
    
    There are multiple y-values for one x-value. Also, class is not a numerical
    variable, but a categorial one.

## Aesthetic mappings

### Content
use another aesthetic to add a third variable, class, and 
use color at unique levels to differenciate each unique value 
of the class-variable

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

other possible aethetics:

*    size  (=> size of the points)
*    alpha (=> opacity)
*    shape (=> shapes, only six possible ones; rest will stay unplotted)

manually setting the aesthetic:
since setting an aesthetic to a constant is independent of any variable, set this **outside** of the aes()-function

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

### Exercises

1.  What's gone wrong with this code? Why are the points not blue?

    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
    ```
    
    It's not blue because the color setting was used inside aes()
    The correct way:
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
    ```

1.  Which variables in `mpg` are categorical? Which variables are continuous? 
    (Hint: type `?mpg` to read the documentation for the dataset). How
    can you see this information when you run `mpg`?
    
    ```{r}
    lapply(mpg, class)
    ```
    
    continuous variables:
    ```{r}
    colnames(mpg)[grep("numeric", lapply(mpg, class))]
    ```
    
    categorial variables:
    ```{r}
    colnames(mpg)[grep("character", lapply(mpg, class))]
    ```
    
    discrete variables:
    ```{r}
    colnames(mpg)[grep("integer", lapply(mpg, class))]    
    ```

1.  Map a continuous variable to `color`, `size`, and `shape`. How do
    these aesthetics behave differently for categorical vs. continuous
    variables? 
    
    `color`:
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, color = displ))
    ```
    
    `size`:
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, size = displ))
    ```
    
    `shape`:
    ```{r, eval = F}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, shape = displ))
    ```
    `size` & `color` work well, `shape` is more suited for categorial variables
    
    `color`:
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, color = drv))
    ```
    
    `size`:
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, size = drv))
    ```
    
    `shape`:
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, shape = drv))
    ```
    
    `color` and `shape` work well, `size` is better for discrete/continuous
    variables
    
1.  What happens if you map the same variable to multiple aesthetics? 
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, color = hwy))
    ```
    
    The plot essentially loses information, since the same thing is displayed
    twice

1.  What does the `stroke` aesthetic do? What shapes does it work with?
    (Hint: use `?geom_point`)
    
    ```{r, eval = F}
    ?geom_point
    ```
    
    > For shapes that have a border (like 21), you can colour the inside and
    > outside separately. Use the stroke aesthetic to modify the width of the
    > border
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, stroke = 2))
    ```
    
1.  What happens if you map an aesthetic to something other than a variable 
    name, like `aes(colour = displ < 5)`? 
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = cty, y = hwy, color = displ < 5))
    ```
    
    This creates a logical vector indicating whether the condition is true or
    not and displays these two levels as the third aesthetic in the legend

## Common problems

## Facets
### Content
### Exercises

1.  What happens if you facet on a continuous variable?

    ```{r}
    ggplot(data = mpg) +
      geom_point(aes(x = drv, y = cyl)) +
      facet_wrap(~ displ, nrow = 2)
    ```
  
    For every realisation of the variable a new facet will be added.
    
1.  What do the empty cells in plot with `facet_grid(drv ~ cyl)` mean?
    How do they relate to this plot?
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = drv, y = cyl))
    ```
    
    There are points missing whenever there is no realisation in the data as
    `cyl` for that `drv`.

1.  What plots does the following code make? What does `.` do?

    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy)) +
      facet_grid(drv ~ .)
    
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy)) +
      facet_grid(. ~ cyl)
    ```
    
    The `facet_grid` formula works according to the pattern `y ~ x`, where the 
    y variable is mapped along the y-axis of the plot (as rows) and the x
    variable is mapped along the x-axis (as columns).
    
    When instead of a variable, there is a dot, that means to omit that
    dimension. (`. ~ x` to map `x` along columns and `y ~ .` to map `y` along
    rows).

1.  Take the first faceted plot in this section:

    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy)) + 
      facet_wrap(~ class, nrow = 2)
    ```
    
    What are the advantages to using faceting instead of the colour aesthetic?
    What are the disadvantages? How might the balance change if you had a 
    larger dataset?
    
    *   advantages:
        Faceting gives a plot per level. That means that it's easier to spot
        how the data for each level is distributed within the x- and y-axes.
        This is especially helpful for larger datasets, where the data points 
        from different facets might overlap.
        
    *   disadvantages:
        Since levels are split up over multiple plots, it is harder to spot 
        outliers or clusters in the data.
    
1.  Read `?facet_wrap`. What does `nrow` do? What does `ncol` do? What other
    options control the layout of the individual panels? Why doesn't
    `facet_grid()` have `nrow` and `ncol` variables?
    
    `facet_wrap()` builds facetting for one vector. `nrow` and `ncol` set the
    number of rows and columns along which the facets of that vector should be 
    distributed in the plot.
    
    `facet_grid()` has predetermined numbers of rows and columns. The number of
    rows is equal to the number of levels for the `y` variable and the number 
    of columns to the number of level for the `x` variable specified in the 
    formula `y ~ x`.

1.  When using `facet_grid()` you should usually put the variable with more
    unique levels in the columns. Why?
    
    This makes sense because most displaying devices have a larger number of 
    pixels on the horizontal than the vertical axis.

## Geometric objects

### Content

### Exercises

1.  What geom would you use to draw a line chart? A boxplot? 
    A histogram? An area chart?
    
    *   line: 
        `geom_line()`
        
    *   boxplot: 
        `geom_boxplot()`
        
    *   histogram:
        `geom_histogram()`
        
    *   area chart:
        `geom_area()`
        
        
1.  Run this code in your head and predict what the output will look like.
    Then, run the code in R and check your predictions.
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
      geom_point() + 
      geom_smooth(se = FALSE)
    ```
    
    This plot will show a scatter plot of `hwy` against `displ` with an 
    additional aesthetic for `drv` by color. On top, there will be a smoothed 
    regression line per level of `drv` without a confidence interval being 
    displayed.

1.  What does `show.legend = FALSE` do?  What happens if you remove it?  
    Why do you think I used it earlier in the chapter?
    
    It didn't occur earlier in the chapter...
    But it stops the legend from being displayed.

1.  What does the `se` argument to `geom_smooth()` do?
    
    `se` shows whether a confidence interval should be displayed around the 
    smoothed curve.

1.  Will these two graphs look different? Why/why not?

    ```{r, eval = FALSE}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point() + 
      geom_smooth()
    
    ggplot() + 
      geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
    ```
    
    Yes, they will. In the first, options are set in the `ggplot()`-function 
    and passed down to the next layers. In the second plot, the options are set
    manually per layer to the same values.

1.  Recreate the R code necessary to generate the following graphs.
    
    ```{r echo = FALSE, fig.width = 3, out.width = "50%", fig.align = "default", message = FALSE}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point() + 
      geom_smooth(se = FALSE)
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_smooth(aes(group = drv), se = FALSE) +
      geom_point()
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
      geom_point() + 
      geom_smooth(se = FALSE)
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point(aes(color = drv)) + 
      geom_smooth(se = FALSE)
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point(aes(color = drv)) +
      geom_smooth(aes(linetype = drv), se = FALSE)
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point(size = 4, colour = "white") + 
      geom_point(aes(colour = drv))
    ```

## Statistical transformations

## Position adjustments

## Coordinate systems

## The layered grammar of graphics

